Oriented Gradient Histogram applied to P300 Detection

### 2017-12-13 article.tex
Subo los cambios de Juliana.  Agregué el detalle de cómo fue la sesión de captura del dataset que tomamos.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradients of the Signal Plot method is explained. Section \ref{Preprocessing} and \ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

@@

@@

On the generated image, a keypoint $\mathbf{kp}$ is placed on a pixel $(x_{kp}, y_{kp})$ over the image plot and a window around the keypoint is considered. A local image patch of size $S_p \times S_p$ is constructed by dividing the window in $16$ blocks of size $3s$ each one,  where $s$ is the scale of the local patch and it is an input parameter of the algorithm. It is arranged in a $4 \times 4$ grid and the pixel $ \mathbf{kp}$ is the patch center, then $S_p = 4.3s $ pixels.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c)
\label{averaging}
\end{equation}

\item Plot the signal $x(t,c)$, according Section~\ref{Plot}.

\item Obtain the descriptor from the image of the signal plot of $x(t,c)$ in accordance to the method described in Section~\ref{SIFT}, $ d^r_1, \dots,  d^r_6 $ for rows and  $ d^c_7, \dots,  d^c_{12} $ for columns.

@@

\begin{equation}
\hat{r} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^r_u)}^{} \left\lVert q -  d^r_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^c_u)}^{} \left\lVert q -  d^c_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l=r,c$ is the set of the $k$ nearest neighbors to $d^l_u, \;l=r,c$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l=r,c$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $.

@@

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}
\centering
\includegraphics[width=15cm]{classificationgraph2.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

@@

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor
\end{equation}

\noindent where $\gamma$ is the image scale, $t$ is time and $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for  channel $c$. Lastly, $\bar{x}(c)$ and $ \sigma (c) $ are the mean and standard deviation of $x$.

\noindent where $ z_1$ and $z_2$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image.  The function $z(c)$ is the \textit{zerolevel} which is the location on the image where the signal's zero value should be located in order to fit the entire signal within the image:

@@

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ in the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds in the horizontal direction. Finally, descriptors  $\mathbf{kp}$  are located at $ x_{kp} =  0.55 Fs.\gamma = 35 $ and $ y_{kp} = z(c) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction). 
% esta figura no está comentada en el texto.
\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole signal information.  The vertical size $4 \cdot 3 s_y$ of the patch is required to be long enough as to contain the signal $\Delta  \mu V $, which is the peak-to-peak amplitude of the transient event. The horizontal size $ 4 \cdot 3 s_x$ , on the other hand, needs to cover the entire duration in seconds of the transient signal event, $ \lambda $.   }
\label{fig:sampledescriptor2}
\end{figure}

@@

Results are shown in Table \ref{tab:results} where the percentage of correctly spelled letters is calculated while performing an offline BCI Simulation for the public dataset of ALS patients.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well.

DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel, more representative of the difference in amplitudes. However, in our approach occipital channels,  PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject. Chance level is $2\%$. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   $35\%$  \\
2     &     Fz   &   $85\%$  \\
3     &     Cz   &   $25\%$  \\
4     &     PO8 &   $55\%$  \\
5     &     PO7 &   $40\%$ \\
6     &     PO7 &   $60\%$  \\
7     &     PO8 &   $80\%$  \\
8     &     PO7 &   $95\%$ \\

@@

The ITR, or BTR, in the case of reactive BCIs \citep{WolpawJonathanR2012} strongly depends on the amount of signal averaging required to transmit a valid and robust selection.  The Performance curves (Fig. \ref{fig:performance}) show how the percentage of correctly identified letters depends on the number of repetitions that were used to obtain the averaged signal.
For both datasets, the experimental protocol uses a very short ISI which has the potential to increase the ITR but at the same time is inversily proportional to the amplitude of the p300 response (i.e. harder to detect).  As an unexpected consequence, 15 out of 20 subject reported high ocular tiredness by using this configuration.

>>>>

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradients of the Signal Plot method is explained. Section~\ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Nearest Neighbor (NBNN)~\citep{Boiman2008} and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

@@

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma(c) } }\right \rfloor, t\geq 0,c \in \{1,2,\dots,8\}
\end{equation}

\noindent where $\gamma > 0$ is an input parameter of the algorithm and  it is related with the image scale. On the other hand, $ x(t,c) $ is the point-to-point averaged EEG matrix for time $t$ and for channel $c$. Lastly, $\bar{x}(c)$ and $ \sigma (c) $ are the mean and standard deviation of $x$.

\noindent where $ (z_1,z_2) \in \mathcal{N} \times \mathcal{N}$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image.  The function $z(c)$ is the location on the image where the signal's zero value has to be located in order to fit the entire signal within the image:

@@

On the generated image $I$, a keypoint $\mathbf{kp}$ is placed on a pixel $(x_{kp}, y_{kp})$ over the image plot and a window around the keypoint is considered. A local image patch of size $S_p \times S_p$ pixels is constructed by dividing the window in $16$ blocks of size $3s$ each one,  where $s$ is the scale of the local patch and it is an input parameter of the algorithm. It is arranged in a $4 \times 4$ grid and the pixel $ \mathbf{kp}$ is the patch center, then $S_p = 4.3s $ pixels.

@@

\begin{enumerate}
\item Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject. For each illuminated row or column, an image is obtained.
\item Repeat item 1 $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t$ and $c$ correspond to time and chanel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t\geq 0, c \in \{1,\dots,8\}
\label{averaging}
\end{equation}  
for each row and column.

\item Plot the signal $x(t,c)$, according Section~\ref{Plot}. The images $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$ are obtained for rows and columns, respectively.

\item Obtain the descriptors $ d^{row}_1, \dots,  d^{row}_6 $  and  $ d^{col}_7, \dots,  d^{col}_{12} $ for rows and columns, respectively from $I^{row}_1, \dots, I^{row}_6$ and $I^{col}_7,\dots,I^{col}_{12}$  in accordance to the method described in Section~\ref{SIFT}.

@@

\begin{equation}
\hat{r} = \arg \min_{u \in \{1,\dots,6\}} \sum_{q \in NN_T(d^{row}_u)}^{} \left\lVert q -  d^{row}_u \right\rVert ^2
\label{eq:multiclassificationrow}
\end{equation}

@@

\begin{equation}
\hat{c} = \arg \min_{u \in \{7,\dots,12\}} \sum_{q \in NN_T(d^{col}_u)}^{} \left\lVert q -  d^{col}_u \right\rVert ^2
\label{eq:multiclassificationcol}
\end{equation}

where $NN_T(d^l_u), \;l\in\{row,{col}\}$ is the set of the $k$ nearest neighbors to $d^l_u, \;l\in\{row,{col}\}$ and $q$ is a template descriptor that belongs to $NN_T(d^l_u), \;l\in\{row,{col}\}$.  By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{r} $ and column $ \hat{c} $.

@@

Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph2.pdf}
\caption{Single trial segments are averaged for the 6 rows and 6 columns.}
\label{fig:classification}
\end{figure}

@@

@@

EEG data is collected in a single recording session. Participants are seated in a comfortable chair, with their vision aligned to a computer screen located one meter in front of them.  The handling and processing of the data and stimulations is handled by the OpenVibe platform.

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ in the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds in the horizontal direction. Finally, descriptors  $\mathbf{kp}$  are located at $ (x_{kp}, y_{kp} )= ( 0.55 Fs.\gamma, z(c) )= (35,  z(c)) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction).

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{patchgeometry.pdf}
\caption{The scale of local patch is selected in order to capture the whole signal information.  The vertical size $4 \cdot 3 s_y$ of the patch is required to be long enough as to contain the signal $\Delta  \mu V $, which is the peak-to-peak amplitude of the transient event. The horizontal size $ 4 \cdot 3 s_x$ , on the other hand, needs to cover the entire duration in seconds of the transient signal event, $ \lambda $.   }
\label{fig:sampledescriptor2}
\end{figure}

@@

Table~\ref{tab:results} shows the results of applying the algorithm to the subjects of the public dataset. The percentage of correctly spelled letters is calculated while performing an offline BCI Simulation for the public dataset of ALS patients.  From the 7 words for each subject, the first 3 were used as training, and the remaining 4 for testing.  The best performing channel is informed as well. The chance level is $2\%$.  
%JG esto después lo hablamos, ¿porque lo pondrías aca?
%DRAFT: Traditionally P300 can be seen in this channel and this other channel.  However, Wolpaw showed that the response may also arise in occipital channels as well.  The authors of the original paper concluded that Cz was the best performing channel, more representative of the difference in amplitudes. However, in our approach occipital channels,  PO8 and PO7 showed higher performances \citep{Huggins2016,Jure2016}.

\begin{table}[H]
\caption{Percentage of correctly predicted letters while performing an offline BCI Simulation for the best performing channel for each subject.  }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Participant}	&  \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &     Cz   &   $35\%$  \\
2     &     Fz   &   $85\%$  \\
3     &     Cz   &   $25\%$  \\
4     &     PO8 &   $55\%$  \\
5     &     PO7 &   $40\%$ \\
6     &     PO7 &   $60\%$  \\
7     &     PO8 &   $80\%$  \\
8     &     PO7 &   $95\%$ \\

@@

The ITR, or BTR, in the case of reactive BCIs~\citep{WolpawJonathanR2012}  depends on the amount of signal averaging required to transmit a valid and robust selection.  Fig.~\ref{fig:performance} shows the performance curves. It can be observed that the percentage of correctly identified letters depends on the number of repetitions that were used to obtain the averaged signal.
For both datasets, the experimental protocol uses a very short ISI which has the potential to increase the ITR but at the same time it is inversely proportional to the amplitude of the p300 response (i.e. harder to detect).  As an unexpected consequence, 15 out of 20 subject reported high ocular tiredness by using this configuration.
